---
title: "SHRS Market Analysis — BLS Projections 2016–2026"
author: "SHRS MQE Capstone Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# 0 — Setup & Data Loading

## 0.1 Packages

```{r packages}
library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)
```

## 0.2 SOC Crosswalk — All 9 SHRS Programs

The 2016 cycle uses SOC 29-2071 for HIM (same as 2018).

```{r crosswalk}
soc_crosswalk <- tribble(
  ~shrs_program, ~shrs_dept,  ~soc_code,  ~occupation_title,
  "SLP",         "CSD",       "29-1127",  "Speech-Language Pathologists",
  "AuD",         "CSD",       "29-1181",  "Audiologists",
  "HIM",         "HIM",       "29-2071",  "Medical Records and Health Information Technicians",
  "OTD",         "RST",       "29-1122",  "Occupational Therapists",
  "DPT",         "RST",       "29-1123",  "Physical Therapists",
  "PAS",         "PAS",       "29-1071",  "Physician Assistants",
  "AT",          "SMN",       "29-9091",  "Athletic Trainers",
  "DN",          "SMN",       "29-1031",  "Dietitians and Nutritionists",
  "SS",          "SMN",       "29-1128",  "Exercise Physiologists"
)

target_socs <- soc_crosswalk$soc_code
```

## 0.3 Load Data

Same 2-row header format as 2018. Skip row 2, assign column names by position.

```{r load-data}
PROJ_CSV_PATH <- file.path(path.expand("~"), "Documents", "SHRS_Analysis",
                            "market_analysis", "v3data", "projections_csv",
                            "occupation_csv", "occupation_2016.csv")

raw <- read_csv(PROJ_CSV_PATH, skip = 1, col_names = FALSE,
                 show_col_types = FALSE)

names(raw) <- c(
  "occ_title", "soc_code", "occ_type",
  "emp_base", "emp_projected",
  "emp_change_num", "emp_change_pct",
  "lf_exit_rate", "occ_transfer_rate", "total_sep_rate",
  "lf_exits", "occ_transfers", "total_separations",
  "annual_openings"
)

raw <- raw |>
  filter(!is.na(soc_code), str_detect(soc_code, "^\\d{2}-\\d{4}$"))

raw <- raw |>
  mutate(across(c(emp_base:annual_openings), ~ {
    x <- str_remove_all(as.character(.x), ",")
    x <- str_trim(x)
    x <- na_if(x, "—")
    x <- na_if(x, "--")
    as.numeric(x)
  }))
```

## 0.4 Benchmarks & SHRS Subset

```{r benchmarks}
all_occ <- raw |> filter(soc_code == "00-0000")
all_occ_growth_pct   <- all_occ$emp_change_pct
all_occ_sep_rate     <- all_occ$total_sep_rate
all_occ_openings     <- all_occ$annual_openings
CYCLE <- "2016-2026"

cat("All-Occupations benchmark (", CYCLE, "):\n")
cat("  Growth:", all_occ_growth_pct, "%\n")
cat("  Sep rate:", all_occ_sep_rate, "%\n")
cat("  Annual openings:", comma(all_occ_openings * 1000), "\n")

proj <- raw |>
  filter(soc_code %in% target_socs,
         str_detect(occ_type, regex("line item", ignore_case = TRUE))) |>
  left_join(soc_crosswalk, by = "soc_code") |>
  select(shrs_program, shrs_dept, soc_code, occ_title = occupation_title,
         emp_base, emp_projected, emp_change_num, emp_change_pct,
         lf_exit_rate, occ_transfer_rate, total_sep_rate,
         lf_exits, occ_transfers, total_separations, annual_openings) |>
  arrange(shrs_dept, shrs_program)

cat("\nSHRS programs loaded:", nrow(proj), "of 9 expected\n")
cat("Programs found:", paste(proj$shrs_program, collapse = ", "), "\n")
```

---

# 1 — Employment Growth Analysis

## 1.1 Projected Growth Rates vs Benchmark

```{r growth-bar, fig.height=6}
base_yr <- str_extract(CYCLE, "^\\d{4}")
proj_yr <- str_extract(CYCLE, "\\d{4}$")

proj |>
  mutate(
    health = case_when(
      emp_change_pct >= all_occ_growth_pct * 2 ~ "Strong (≥2× avg)",
      emp_change_pct >= all_occ_growth_pct     ~ "Healthy (≥ avg)",
      emp_change_pct >= 0                       ~ "Caution (below avg)",
      TRUE                                      ~ "Declining"
    ),
    health = factor(health, levels = c("Strong (≥2× avg)", "Healthy (≥ avg)",
                                        "Caution (below avg)", "Declining"))
  ) |>
  ggplot(aes(x = reorder(shrs_program, emp_change_pct), y = emp_change_pct,
             fill = health)) +
  geom_col(width = 0.65) +
  geom_hline(yintercept = all_occ_growth_pct, linetype = "dashed",
             color = "gray30", linewidth = 0.8) +
  geom_text(aes(label = paste0(emp_change_pct, "%")),
            hjust = -0.15, size = 3.8, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c(
    "Strong (≥2× avg)"   = "#1a9850",
    "Healthy (≥ avg)"     = "#91cf60",
    "Caution (below avg)" = "#fee08b",
    "Declining"           = "#d73027"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  annotate("text", x = 0.5, y = all_occ_growth_pct,
           label = paste0("All occupations: ", all_occ_growth_pct, "%"),
           hjust = -0.05, vjust = -0.5, size = 3.3, color = "gray30") +
  labs(
    title    = paste0("Projected Employment Growth by SHRS Program (", CYCLE, ")"),
    subtitle = "BLS National Employment Matrix | Earliest cycle in this analysis",
    x = NULL, y = "Projected Growth (%)", fill = "Market Signal"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## 1.2 Employment Levels

```{r emp-levels, fig.height=6}
proj |>
  select(shrs_program, emp_base, emp_projected) |>
  pivot_longer(cols = c(emp_base, emp_projected), names_to = "period",
               values_to = "employment") |>
  mutate(period = if_else(period == "emp_base", base_yr, proj_yr)) |>
  ggplot(aes(x = reorder(shrs_program, employment), y = employment,
             fill = period)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = comma(employment, accuracy = 0.1)),
            position = position_dodge(width = 0.6), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(values = setNames(c("#4393c3", "#2166ac"),
                                       c(base_yr, proj_yr))) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.25))) +
  labs(title = paste0("Employment: ", base_yr, " vs ", proj_yr, " (thousands)"),
       x = NULL, y = "Employment (thousands)", fill = "Year") +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

## 1.3 Growth Summary Table

```{r growth-table}
proj |>
  mutate(
    vs_benchmark_pp = round(emp_change_pct - all_occ_growth_pct, 1),
    growth_signal = case_when(
      emp_change_pct >= all_occ_growth_pct * 2 ~ "🟢 Strong",
      emp_change_pct >= all_occ_growth_pct     ~ "🟢 Healthy",
      emp_change_pct >= 0                       ~ "🟡 Caution",
      TRUE                                      ~ "🔴 Declining"
    )
  ) |>
  select(Program = shrs_program, Dept = shrs_dept,
         `Emp Base (K)` = emp_base, `Emp Proj (K)` = emp_projected,
         `Change (K)` = emp_change_num, `Growth (%)` = emp_change_pct,
         `vs Bench (pp)` = vs_benchmark_pp, Signal = growth_signal) |>
  arrange(desc(`Growth (%)`)) |>
  kable(format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 2 — Job Openings & Replacement Demand

## 2.1 Components of Annual Openings

```{r openings-components, fig.height=6}
proj |>
  mutate(growth_component = pmax(emp_change_num / 10, 0)) |>
  select(shrs_program, `Labor Force Exits` = lf_exits,
         `Occupational Transfers` = occ_transfers,
         `Growth` = growth_component) |>
  pivot_longer(-shrs_program, names_to = "component", values_to = "value") |>
  mutate(component = factor(component,
                             levels = c("Growth", "Occupational Transfers",
                                        "Labor Force Exits"))) |>
  ggplot(aes(x = reorder(shrs_program, value, sum), y = value,
             fill = component)) +
  geom_col(width = 0.6) + coord_flip() +
  scale_fill_manual(values = c("Growth" = "#2166ac",
                                "Occupational Transfers" = "#fee08b",
                                "Labor Force Exits" = "#d73027")) +
  labs(title = paste0("Components of Annual Job Openings (", CYCLE, ")"),
       x = NULL, y = "Annual Openings (thousands)", fill = NULL) +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

## 2.2 Separation Rates

```{r sep-rates, fig.height=6}
proj |>
  select(shrs_program, lf_exit_rate, occ_transfer_rate) |>
  pivot_longer(-shrs_program, names_to = "type", values_to = "rate") |>
  mutate(type = if_else(type == "lf_exit_rate", "Labor Force Exit Rate",
                        "Occupational Transfer Rate")) |>
  ggplot(aes(x = reorder(shrs_program, rate, sum), y = rate, fill = type)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = all_occ_sep_rate, linetype = "dashed",
             color = "gray30") +
  coord_flip() +
  scale_fill_manual(values = c("Labor Force Exit Rate" = "#d73027",
                                "Occupational Transfer Rate" = "#fee08b")) +
  labs(title = paste0("Separation Rates (", CYCLE, ")"),
       x = NULL, y = "Rate (%)", fill = NULL) +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

## 2.3 Openings Table

```{r openings-table}
proj |>
  mutate(
    openings_signal = case_when(
      annual_openings >= 10 ~ "🟢 High Volume", annual_openings >= 3 ~ "🟢 Solid",
      annual_openings >= 1 ~ "🟡 Moderate", TRUE ~ "🔴 Limited"),
    replacement_signal = case_when(
      total_sep_rate >= 6 ~ "🟢 High Replacement",
      total_sep_rate >= 4 ~ "🟡 Moderate", TRUE ~ "🔴 Low")
  ) |>
  select(Program = shrs_program, `Exit Rate (%)` = lf_exit_rate,
         `Transfer Rate (%)` = occ_transfer_rate,
         `Total Sep Rate (%)` = total_sep_rate,
         `Openings/yr (K)` = annual_openings,
         `Openings Signal` = openings_signal,
         `Replacement Signal` = replacement_signal) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 3 — Market Size & Scale

```{r scale-scatter, fig.height=6}
proj |>
  mutate(market_size = case_when(
    emp_base >= 200 ~ "Large (≥200K)", emp_base >= 50 ~ "Mid-size (50–200K)",
    emp_base >= 20 ~ "Small (20–50K)", TRUE ~ "Niche (<20K)"),
    market_size = factor(market_size, levels = c("Large (≥200K)",
      "Mid-size (50–200K)", "Small (20–50K)", "Niche (<20K)"))) |>
  ggplot(aes(x = emp_base, y = emp_change_pct, color = market_size,
             label = shrs_program)) +
  geom_point(size = 5) +
  geom_text(vjust = -1.2, size = 4, fontface = "bold") +
  geom_hline(yintercept = all_occ_growth_pct, linetype = "dashed",
             color = "gray40") +
  scale_color_manual(values = c("Large (≥200K)" = "#1a9850",
    "Mid-size (50–200K)" = "#4393c3", "Small (20–50K)" = "#fee08b",
    "Niche (<20K)" = "#d73027")) +
  scale_x_continuous(labels = comma) +
  labs(title = paste0("Market Size vs Growth (", CYCLE, ")"),
       x = paste0(base_yr, " Employment (thousands)"),
       y = "Projected Growth (%)", color = "Market Size") +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

```{r openings-intensity}
proj |>
  mutate(openings_intensity = round(annual_openings / emp_base * 1000, 1)) |>
  ggplot(aes(x = reorder(shrs_program, openings_intensity),
             y = openings_intensity,
             fill = if_else(openings_intensity >= 70, "High", "Normal"))) +
  geom_col(width = 0.6) +
  geom_text(aes(label = openings_intensity), hjust = -0.15, size = 3.8) +
  coord_flip() +
  scale_fill_manual(values = c("High" = "#1a9850", "Normal" = "#4393c3"),
                     guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(title = "Openings Intensity (per 1,000 workers)", subtitle = CYCLE,
       x = NULL, y = "Openings per 1,000 workers") +
  theme_minimal(base_size = 13)
```

---

# 4 — Scorecard (0–100 Scale)

```{r scorecard-compute}
rescale_0_100 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (rng[2] == rng[1]) return(rep(50, length(x)))
  round((x - rng[1]) / (rng[2] - rng[1]) * 100, 1)
}

scorecard <- proj |>
  mutate(
    growth_score    = rescale_0_100(emp_change_pct),
    openings_score  = rescale_0_100(annual_openings),
    sep_rate_score  = rescale_0_100(total_sep_rate),
    emp_scale_score = rescale_0_100(emp_base),
    composite_score = round((growth_score + openings_score +
                              sep_rate_score + emp_scale_score) / 4, 1),
    health_signal = case_when(
      composite_score >= 65 ~ "Green", composite_score >= 40 ~ "Yellow",
      TRUE ~ "Red"),
    health_signal = factor(health_signal, levels = c("Green", "Yellow", "Red"))
  )
```

```{r scorecard-heatmap, fig.height=6, fig.width=10}
scorecard |>
  select(shrs_program, growth_score, openings_score,
         sep_rate_score, emp_scale_score) |>
  pivot_longer(-shrs_program, names_to = "dimension", values_to = "score") |>
  mutate(dimension = case_when(
    dimension == "growth_score" ~ "Projected Growth",
    dimension == "openings_score" ~ "Annual Openings",
    dimension == "sep_rate_score" ~ "Replacement Demand",
    dimension == "emp_scale_score" ~ "Market Size")) |>
  ggplot(aes(x = shrs_program, y = dimension, fill = score)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = score), size = 4, fontface = "bold") +
  scale_fill_gradient2(low = "#d73027", mid = "#fee08b", high = "#1a9850",
                        midpoint = 50, limits = c(0, 100),
                        name = "Score\n(0–100)") +
  labs(title = paste0("Market Demand Scorecard (", CYCLE, ")"),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(face = "bold", angle = 30, hjust = 1),
        panel.grid = element_blank())
```

```{r composite-bar, fig.height=5}
scorecard |>
  ggplot(aes(x = reorder(shrs_program, composite_score), y = composite_score,
             fill = health_signal)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = composite_score), hjust = -0.15, size = 4,
            fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c("Green" = "#1a9850", "Yellow" = "#fee08b",
                                "Red" = "#d73027")) +
  scale_y_continuous(limits = c(0, 105)) +
  labs(title = paste0("Composite Market Score (", CYCLE, ")"),
       x = NULL, y = "Composite Score (0–100)", fill = "Health") +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

```{r scorecard-table}
scorecard |>
  select(Program = shrs_program, Dept = shrs_dept,
         `Growth (%)` = emp_change_pct, `Openings/yr (K)` = annual_openings,
         `Sep Rate (%)` = total_sep_rate, `Emp Base (K)` = emp_base,
         `Growth Sc` = growth_score, `Open Sc` = openings_score,
         `Sep Sc` = sep_rate_score, `Scale Sc` = emp_scale_score,
         Composite = composite_score, Signal = health_signal) |>
  arrange(desc(Composite)) |>
  kable(format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 5 — Statistics

```{r stats-summary}
cat("===", CYCLE, "===\n")
cat("Growth: Mean:", round(mean(proj$emp_change_pct), 2),
    "| Median:", round(median(proj$emp_change_pct), 2),
    "| SD:", round(sd(proj$emp_change_pct), 2), "\n")
cat("Openings: Mean:", round(mean(proj$annual_openings), 2),
    "| Total:", round(sum(proj$annual_openings), 2), "K\n")
cat("Sep Rate: Mean:", round(mean(proj$total_sep_rate), 2), "%\n")
cat("Above benchmark:", sum(proj$emp_change_pct >= all_occ_growth_pct), "of 9\n")
```

```{r corr}
cor_growth_openings <- cor(proj$emp_change_pct, proj$annual_openings,
                            use = "complete.obs")
cor_growth_sep <- cor(proj$emp_change_pct, proj$total_sep_rate,
                       use = "complete.obs")
cat("Corr Growth↔Openings:", round(cor_growth_openings, 3), "\n")
cat("Corr Growth↔Sep:", round(cor_growth_sep, 3), "\n")
```

---

# 6 — Stored Outputs

```{r stored-outputs}
proj_2016_data       <- proj
scorecard_2016_data  <- scorecard
benchmark_2016 <- list(cycle = CYCLE, all_occ_growth = all_occ_growth_pct,
                        all_occ_sep_rate = all_occ_sep_rate,
                        all_occ_openings = all_occ_openings)
summary_stats_2016 <- list(
  growth_mean = round(mean(proj$emp_change_pct), 2),
  growth_median = round(median(proj$emp_change_pct), 2),
  growth_sd = round(sd(proj$emp_change_pct), 2),
  openings_total = round(sum(proj$annual_openings), 2),
  openings_mean = round(mean(proj$annual_openings), 2),
  sep_rate_mean = round(mean(proj$total_sep_rate), 2),
  n_above_bench = sum(proj$emp_change_pct >= all_occ_growth_pct),
  n_below_bench = sum(proj$emp_change_pct < all_occ_growth_pct),
  cor_growth_open = round(cor_growth_openings, 3),
  cor_growth_sep = round(cor_growth_sep, 3))
program_signals_2016 <- scorecard |>
  select(shrs_program, shrs_dept, composite_score, health_signal,
         emp_change_pct, annual_openings, total_sep_rate, emp_base)
soc_crosswalk_9 <- soc_crosswalk

cat("=== Stored Objects (", CYCLE, ") ===\n")
cat("  proj_2016_data, scorecard_2016_data, benchmark_2016\n")
cat("  summary_stats_2016, program_signals_2016, soc_crosswalk_9\n")
cat("  HIM SOC: 29-2071 (predecessor code)\n")
```

---

# Appendix

**Source:** BLS National Employment Matrix, 2016–2026 (`occupation_2016.csv`)

**Context:** This is the **earliest cycle** in our 8-cycle analysis series.
It provides the longest historical baseline for tracking how BLS projections
have evolved over nearly a decade of successive revisions.

**File format:** 2-row header (skip row 2).
**HIM SOC:** 29-2071.
**Scorecard:** 0–100, 4 dimensions, Green/Yellow/Red.

**All 9 programs:** SLP, AuD, HIM, OTD, DPT, PAS, AT, DN, SS
