---
title: "SHRS Market Analysis — Detailed Industry Employment (2024 Vintage)"
subtitle: "BLS Employment Projections Table 2.11 · 2024–2034 Cycle"
author: "SHRS MQE Capstone Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# 0 — Setup

```{r packages}
library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)
```

```{r config}
soc_crosswalk <- tribble(
  ~shrs_program, ~shrs_dept, ~soc_code,  ~occupation_title,
  "SLP",         "CSD",      "29-1127",  "Speech-Language Pathologists",
  "AuD",         "CSD",      "29-1181",  "Audiologists",
  "HIM",         "HIM",      "29-9021",  "Health Information Technologists and Medical Registrars",
  "OTD",         "RST",      "29-1122",  "Occupational Therapists",
  "DPT",         "RST",      "29-1123",  "Physical Therapists",
  "PAS",         "PAS",      "29-1071",  "Physician Assistants",
  "AT",          "SMN",      "29-9091",  "Athletic Trainers",
  "DN",          "SMN",      "29-1031",  "Dietitians and Nutritionists",
  "SS",          "SMN",      "29-1128",  "Exercise Physiologists"
)

DATA_PATH <- "/Users/gpps/Documents/SHRS_Analysis/market_analysis/v3data/projections_csv/industry_csv/2024/2024_industry - Table 2.11.csv"

VINTAGE <- 2024
CYCLE   <- "2024-2034"

shrs_naics <- tribble(
  ~naics_code, ~naics_label,                                                    ~shrs_relevance,
  "620000",    "Healthcare and social assistance",                              "Umbrella sector",
  "621000",    "Ambulatory healthcare services",                                "Primary outpatient umbrella",
  "621100",    "Offices of physicians",                                         "PAS, DN setting",
  "621300",    "Offices of other health practitioners",                         "Therapy/audiology umbrella",
  "621340",    "Offices of PT, OT, SLP, and audiologists",                      "Primary for SLP, AuD, OTD, DPT",
  "621390",    "Offices of all other health practitioners",                     "AT, SS, HIM potential",
  "621400",    "Outpatient care centers",                                       "Multi-program setting",
  "621600",    "Home healthcare services",                                      "SLP, OTD, DPT, DN home health",
  "621900",    "Other ambulatory healthcare services",                          "AT, SS umbrella",
  "621990",    "All other ambulatory healthcare services",                      "AT, SS primary",
  "622000",    "Hospitals",                                                     "All SHRS programs",
  "623000",    "Nursing and residential care facilities",                       "SLP, OTD, DPT, DN — SNF/LTC",
  "624000",    "Social assistance",                                             "SLP (early intervention), DN"
)
```

```{r helpers}
clean_num <- function(x) {
  x <- str_remove_all(x, ",")
  x <- str_replace_all(x, "\u2014|\u2013|--|-", NA_character_)
  as.numeric(x)
}
```

---

# 1 — Load & Clean Data

Table 2.11 provides **detailed industry employment** at the 4-digit and 6-digit
NAICS level — the most granular industry data from BLS projections.

**Column layout (2024 vintage, 11 columns):**

| Pos | Content |
|-----|---------|
| 1 | Industry title |
| 2 | National Employment Matrix code |
| 3 | Industry type (Summary / Line item) |
| 4 | Employment, 2024 (thousands) |
| 5 | Employment, 2034 (thousands) |
| 6 | Employment change, numeric, 2024–34 |
| 7 | Employment change, percent, 2024–34 |
| 8 | CAGR, employment |
| 9 | Output, 2024 (billions $) |
| 10 | Output, 2034 (billions $) |
| 11 | CAGR, output |

```{r load-data}
raw <- read_csv(DATA_PATH, col_types = cols(.default = "c"), col_names = FALSE)

colnames(raw) <- c(
  "industry_title", "matrix_code", "industry_type",
  "emp_base", "emp_proj",
  "emp_change", "emp_pct_change", "emp_cagr",
  "output_base", "output_proj", "output_cagr"
)

raw <- raw[-1, ]

raw <- raw %>%
  mutate(industry_title = str_remove_all(industry_title, "\\(\\d+\\)|\\[\\d+\\]") %>%
           str_trim())

df <- raw %>%
  mutate(across(-c(industry_title, matrix_code, industry_type), clean_num))

cat("Loaded", nrow(df), "detailed industry rows for", CYCLE, "cycle\n")
```

```{r preview}
df %>%
  head(10) %>%
  select(industry_title, matrix_code, emp_base, emp_proj, emp_pct_change) %>%
  kable(digits = 1, format.args = list(big.mark = ","),
        caption = "Table 2.11 Preview") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 2 — Benchmarks

```{r benchmarks}
total_row <- df %>% filter(str_detect(matrix_code, "^TE1000")) %>% slice(1)
total_emp      <- total_row$emp_base
total_pct      <- total_row$emp_pct_change
total_cagr     <- total_row$emp_cagr

hc_row <- df %>% filter(matrix_code == "620000") %>% slice(1)
hc_emp_base    <- hc_row$emp_base
hc_emp_proj    <- hc_row$emp_proj
hc_pct         <- hc_row$emp_pct_change
hc_cagr        <- hc_row$emp_cagr

cat("Benchmarks (", CYCLE, "):\n")
cat("  Total economy growth:   ", total_pct, "% | CAGR ", total_cagr, "%\n")
cat("  HC & Social Assistance: ", hc_pct, "% | CAGR ", hc_cagr, "%\n")
cat("  HC/Total ratio:         ", round(hc_pct / total_pct, 2), "x\n")
```

---

# 3 — SHRS-Relevant Industries

## 3.1 Extract & Table

```{r shrs-industries}
shrs_ind <- df %>%
  filter(matrix_code %in% shrs_naics$naics_code) %>%
  left_join(shrs_naics, by = c("matrix_code" = "naics_code")) %>%
  mutate(
    growth_vs_total = round(emp_pct_change - total_pct, 1),
    growth_vs_hc    = round(emp_pct_change - hc_pct, 1)
  )

cat("Found", nrow(shrs_ind), "of", nrow(shrs_naics), "target SHRS industries\n")
```

```{r shrs-table}
shrs_ind %>%
  select(
    Industry         = industry_title,
    Code             = matrix_code,
    `Emp 2024 (K)`   = emp_base,
    `Emp 2034 (K)`   = emp_proj,
    `Change (K)`     = emp_change,
    `Growth (%)`     = emp_pct_change,
    `CAGR (%)`       = emp_cagr,
    `vs Total (pp)`  = growth_vs_total,
    `vs HC (pp)`     = growth_vs_hc
  ) %>%
  kable(digits = 1, format.args = list(big.mark = ","),
        caption = paste0("SHRS-Relevant Industries (", CYCLE, ")")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 3.2 Growth Chart

```{r growth-chart, fig.height=8}
shrs_ind %>%
  filter(!is.na(emp_pct_change)) %>%
  mutate(
    short_name = case_when(
      matrix_code == "620000" ~ "HC & Social Assist (total)",
      matrix_code == "621000" ~ "Ambulatory HC (total)",
      matrix_code == "621100" ~ "Physician offices",
      matrix_code == "621300" ~ "Other health practitioners",
      matrix_code == "621340" ~ "PT/OT/SLP/AuD offices",
      matrix_code == "621390" ~ "All other practitioners",
      matrix_code == "621400" ~ "Outpatient care centers",
      matrix_code == "621600" ~ "Home health services",
      matrix_code == "621900" ~ "Other ambulatory HC",
      matrix_code == "621990" ~ "All other ambulatory HC",
      matrix_code == "622000" ~ "Hospitals",
      matrix_code == "623000" ~ "Nursing & residential care",
      matrix_code == "624000" ~ "Social assistance",
      TRUE ~ str_trunc(industry_title, 35)
    ),
    is_primary = matrix_code %in% c("621340", "621990")
  ) %>%
  ggplot(aes(x = reorder(short_name, emp_pct_change), y = emp_pct_change,
             fill = is_primary)) +
  geom_col(width = 0.65) +
  geom_hline(yintercept = total_pct, linetype = "dashed",
             color = "gray40", linewidth = 0.6) +
  geom_hline(yintercept = hc_pct, linetype = "dotted",
             color = "#d7191c", linewidth = 0.6) +
  annotate("text", x = 2, y = total_pct + 0.3,
           label = paste0("All industries: ", total_pct, "%"),
           hjust = 0, size = 3, color = "gray40") +
  annotate("text", x = 1, y = hc_pct + 0.3,
           label = paste0("HC sector: ", hc_pct, "%"),
           hjust = 0, size = 3, color = "#d7191c") +
  geom_text(aes(label = paste0(emp_pct_change, "%")), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "#2c7bb6", "TRUE" = "#d7191c"),
                    labels = c("Other HC industries", "Primary SHRS industries"),
                    name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
  labs(
    title = paste0("Projected Growth in SHRS-Relevant Industries (", CYCLE, ")"),
    subtitle = "Red = primary SHRS employment industries (621340, 621990)",
    x = NULL, y = "Projected Growth (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## 3.3 Employment Size

```{r size-chart, fig.height=7}
shrs_ind %>%
  filter(!is.na(emp_base), !matrix_code %in% c("620000", "621000")) %>%
  mutate(
    short_name = case_when(
      matrix_code == "621100" ~ "Physician offices",
      matrix_code == "621300" ~ "Other health practitioners",
      matrix_code == "621340" ~ "PT/OT/SLP/AuD offices",
      matrix_code == "621390" ~ "All other practitioners",
      matrix_code == "621400" ~ "Outpatient care centers",
      matrix_code == "621600" ~ "Home health services",
      matrix_code == "621900" ~ "Other ambulatory HC",
      matrix_code == "621990" ~ "All other ambulatory HC",
      matrix_code == "622000" ~ "Hospitals",
      matrix_code == "623000" ~ "Nursing & residential",
      matrix_code == "624000" ~ "Social assistance",
      TRUE ~ str_trunc(industry_title, 30)
    )
  ) %>%
  ggplot(aes(x = reorder(short_name, emp_base), y = emp_base)) +
  geom_col(width = 0.6, fill = "#2c7bb6") +
  geom_text(aes(label = comma(emp_base, accuracy = 0.1)),
            hjust = -0.05, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = paste0("Employment Size in SHRS Industries (", VINTAGE, ", thousands)"),
    x = NULL, y = "Employment (thousands)"
  ) +
  theme_minimal(base_size = 12)
```

---

# 4 — Primary SHRS Industry Deep-Dive

## 4.1 NAICS 621340: PT/OT/SLP/AuD Offices

Primary industry for SLP, AuD, OTD, DPT.

```{r ind-621340}
ind_621340 <- df %>% filter(matrix_code == "621340")

if (nrow(ind_621340) > 0) {
  cat("NAICS 621340\n")
  cat("  Emp (", VINTAGE, "):    ", comma(ind_621340$emp_base * 1000), "\n")
  cat("  Proj (", VINTAGE+10, "):  ", comma(ind_621340$emp_proj * 1000), "\n")
  cat("  Growth:            ", ind_621340$emp_pct_change, "%\n")
  cat("  CAGR:              ", ind_621340$emp_cagr, "%\n")
  cat("  vs Total:          +", round(ind_621340$emp_pct_change - total_pct, 1), " pp\n")
  cat("  vs HC sector:      +", round(ind_621340$emp_pct_change - hc_pct, 1), " pp\n")
}
```

## 4.2 NAICS 621990: All Other Ambulatory HC

Primary industry for AT and SS.

```{r ind-621990}
ind_621990 <- df %>% filter(matrix_code == "621990")

if (nrow(ind_621990) > 0) {
  cat("NAICS 621990\n")
  cat("  Emp (", VINTAGE, "):    ", comma(ind_621990$emp_base * 1000), "\n")
  cat("  Proj (", VINTAGE+10, "):  ", comma(ind_621990$emp_proj * 1000), "\n")
  cat("  Growth:            ", ind_621990$emp_pct_change, "%\n")
  cat("  CAGR:              ", ind_621990$emp_cagr, "%\n")
}
```

## 4.3 Output (Revenue) Trends

```{r output-analysis}
output_ind <- shrs_ind %>%
  filter(!is.na(output_base), !is.na(output_proj),
         !matrix_code %in% c("620000")) %>%
  mutate(
    output_change = output_proj - output_base,
    output_pct    = round((output_change / output_base) * 100, 1),
    short_name = case_when(
      matrix_code == "621000" ~ "Ambulatory HC",
      matrix_code == "621100" ~ "Physician offices",
      matrix_code == "621400" ~ "Outpatient centers",
      matrix_code == "621600" ~ "Home health",
      matrix_code == "623000" ~ "Nursing & residential",
      matrix_code == "624000" ~ "Social assistance",
      TRUE ~ str_trunc(industry_title, 25)
    )
  )

if (nrow(output_ind) > 0) {
  output_ind %>%
    ggplot(aes(x = reorder(short_name, output_cagr), y = output_cagr)) +
    geom_col(width = 0.6, fill = "#1a9641") +
    geom_text(aes(label = paste0(output_cagr, "%")), hjust = -0.1, size = 3.5) +
    coord_flip() +
    labs(
      title = paste0("Output (Revenue) CAGR for SHRS Industries (", CYCLE, ")"),
      subtitle = "Compound annual growth rate of industry output (billions $)",
      x = NULL, y = "Output CAGR (%)"
    ) +
    theme_minimal(base_size = 12)
}
```

---

# 5 — HC Sub-Sector Composition

```{r hc-composition, fig.height=6}
hc_sub <- df %>%
  filter(matrix_code %in% c("621000", "622000", "623000", "624000")) %>%
  mutate(subsector = case_when(
    matrix_code == "621000" ~ "Ambulatory HC",
    matrix_code == "622000" ~ "Hospitals",
    matrix_code == "623000" ~ "Nursing & Residential",
    matrix_code == "624000" ~ "Social Assistance"
  ))

hc_sub %>%
  select(subsector, emp_base, emp_proj) %>%
  pivot_longer(-subsector, names_to = "period", values_to = "emp") %>%
  mutate(period = if_else(period == "emp_base",
                          as.character(VINTAGE), as.character(VINTAGE + 10))) %>%
  ggplot(aes(x = period, y = emp, fill = subsector)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = comma(emp, accuracy = 0.1)),
            position = position_stack(vjust = 0.5), size = 3.5, color = "white") +
  scale_fill_manual(values = c("Ambulatory HC" = "#2c7bb6",
                                "Hospitals" = "#abd9e9",
                                "Nursing & Residential" = "#fdae61",
                                "Social Assistance" = "#d7191c")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = paste0("HC Sub-Sector Composition (", CYCLE, ")"),
    x = NULL, y = "Employment (thousands)", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

hc_sub %>%
  ggplot(aes(x = reorder(subsector, emp_pct_change), y = emp_pct_change,
             fill = subsector)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(emp_pct_change, "%")), hjust = -0.1, size = 4) +
  geom_hline(yintercept = hc_pct, linetype = "dashed", color = "gray50") +
  coord_flip() +
  scale_fill_manual(values = c("Ambulatory HC" = "#2c7bb6",
                                "Hospitals" = "#abd9e9",
                                "Nursing & Residential" = "#fdae61",
                                "Social Assistance" = "#d7191c")) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
  labs(
    title = paste0("HC Sub-Sector Growth Rates (", CYCLE, ")"),
    subtitle = paste0("Dashed = total HC (", hc_pct, "%)"),
    x = NULL, y = "Projected Growth (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

# 6 — Program-to-Industry Scorecard (0–100)

```{r program-map}
program_map <- tribble(
  ~shrs_program, ~primary_naics, ~primary_label,
  "SLP",         "621340",       "PT/OT/SLP/AuD offices",
  "AuD",         "621340",       "PT/OT/SLP/AuD offices",
  "HIM",         "622000",       "Hospitals",
  "OTD",         "621340",       "PT/OT/SLP/AuD offices",
  "DPT",         "621340",       "PT/OT/SLP/AuD offices",
  "PAS",         "621100",       "Physician offices",
  "AT",          "621990",       "All other ambulatory HC",
  "DN",          "621100",       "Physician offices",
  "SS",          "621990",       "All other ambulatory HC"
)

program_ind <- program_map %>%
  left_join(
    df %>% select(matrix_code, emp_base, emp_proj, emp_change,
                  emp_pct_change, emp_cagr),
    by = c("primary_naics" = "matrix_code")
  )
```

```{r industry-scorecard}
program_ind <- program_ind %>%
  mutate(
    growth_score = case_when(
      emp_pct_change >= hc_pct * 2    ~ 100,
      emp_pct_change >= hc_pct * 1.5  ~ 90,
      emp_pct_change >= hc_pct        ~ 80,
      emp_pct_change >= total_pct * 2 ~ 70,
      emp_pct_change >= total_pct     ~ 60,
      emp_pct_change >= 0             ~ 40,
      TRUE                            ~ 20
    ),
    size_score = case_when(
      emp_base >= 3000 ~ 100,
      emp_base >= 1000 ~ 85,
      emp_base >= 500  ~ 70,
      emp_base >= 200  ~ 55,
      emp_base >= 100  ~ 40,
      TRUE             ~ 25
    ),
    cagr_score = case_when(
      emp_cagr >= 2.0 ~ 100,
      emp_cagr >= 1.5 ~ 85,
      emp_cagr >= 1.0 ~ 70,
      emp_cagr >= 0.5 ~ 55,
      emp_cagr >= 0.0 ~ 40,
      TRUE            ~ 20
    ),
    composite_score = round((growth_score + size_score + cagr_score) / 3, 0),
    signal = case_when(
      composite_score >= 70 ~ "Green",
      composite_score >= 50 ~ "Yellow",
      TRUE                  ~ "Red"
    )
  )
```

```{r scorecard-table}
program_ind %>%
  arrange(desc(composite_score)) %>%
  select(
    Program        = shrs_program,
    Industry       = primary_label,
    `Emp (K)`      = emp_base,
    `Growth (%)`   = emp_pct_change,
    `CAGR (%)`     = emp_cagr,
    `Gr Score`     = growth_score,
    `Sz Score`     = size_score,
    `CAGR Score`   = cagr_score,
    Composite      = composite_score,
    Signal         = signal
  ) %>%
  kable(digits = 1, format.args = list(big.mark = ","),
        caption = paste0("Program Industry Health Scorecard (", CYCLE, ")")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(10, bold = TRUE)
```

```{r scorecard-heatmap, fig.height=5}
program_ind %>%
  select(shrs_program, growth_score, size_score, cagr_score) %>%
  pivot_longer(-shrs_program, names_to = "dimension", values_to = "score") %>%
  mutate(dimension = case_when(
    dimension == "growth_score" ~ "Industry Growth",
    dimension == "size_score"   ~ "Industry Size",
    dimension == "cagr_score"   ~ "Industry CAGR"
  )) %>%
  ggplot(aes(x = shrs_program, y = dimension, fill = score)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = score), size = 4, fontface = "bold") +
  scale_fill_gradient2(low = "#dc3545", mid = "#ffc107", high = "#28a745",
                        midpoint = 60, limits = c(0, 100),
                        name = "Score\n(0-100)") +
  labs(
    title = paste0("Industry Health Heatmap by SHRS Program (", CYCLE, ")"),
    subtitle = "Green >= 70 | Yellow 50-69 | Red < 50",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(face = "bold"), panel.grid = element_blank())
```

```{r composite-bar, fig.height=5}
program_ind %>%
  ggplot(aes(x = reorder(shrs_program, composite_score), y = composite_score,
             fill = signal)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = composite_score), hjust = -0.2, size = 4.5, fontface = "bold") +
  geom_hline(yintercept = 70, linetype = "dashed", color = "#155724", linewidth = 0.5) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "#856404", linewidth = 0.5) +
  coord_flip() +
  scale_fill_manual(values = c("Green" = "#28a745", "Yellow" = "#ffc107",
                                "Red" = "#dc3545")) +
  scale_y_continuous(limits = c(0, 110)) +
  labs(
    title = paste0("Industry Composite Score by SHRS Program (", CYCLE, ")"),
    subtitle = "Average of Growth, Size, and CAGR scores (0-100)",
    x = NULL, y = "Composite Score (0-100)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

# 7 — Key Findings

```{r findings, results='asis'}
fastest <- shrs_ind %>%
  filter(!matrix_code %in% c("620000", "621000")) %>%
  slice_max(emp_pct_change, n = 1)
largest <- shrs_ind %>%
  filter(!matrix_code %in% c("620000", "621000")) %>%
  slice_max(emp_base, n = 1)

cat("\n## Industry Environment Summary\n\n")
cat(paste0("- **Fastest-growing SHRS industry**: ", fastest$industry_title,
           " (", fastest$emp_pct_change, "%)\n"))
cat(paste0("- **Largest SHRS employer**: ", largest$industry_title,
           " (", comma(largest$emp_base), "K)\n"))
cat(paste0("- **621340 (PT/OT/SLP/AuD)**: ",
           ind_621340$emp_pct_change, "% growth — ",
           round(ind_621340$emp_pct_change / total_pct, 1), "x total economy\n"))
cat(paste0("- **621990 (Other ambulatory HC)**: ",
           ind_621990$emp_pct_change, "% growth\n"))
cat(paste0("- **HC sector overall**: ", hc_pct, "% growth (",
           round(hc_pct / total_pct, 1), "x total economy)\n"))
```

---

# 8 — Stored Outputs

```{r stored-outputs}
ind_t211_2024_vintage       <- VINTAGE
ind_t211_2024_cycle         <- CYCLE
ind_t211_2024_total_pct     <- total_pct
ind_t211_2024_total_cagr    <- total_cagr
ind_t211_2024_hc_pct        <- hc_pct
ind_t211_2024_hc_cagr       <- hc_cagr
ind_t211_2024_hc_emp        <- hc_emp_base
ind_t211_2024_shrs_ind      <- shrs_ind
ind_t211_2024_ind_621340    <- ind_621340
ind_t211_2024_ind_621990    <- ind_621990
ind_t211_2024_program_scores <- program_ind
ind_t211_2024_hc_subsectors  <- hc_sub
ind_t211_2024_full_df        <- df
```

```{r output-registry}
output_registry <- tribble(
  ~variable,                          ~description,
  "ind_t211_2024_vintage",            "Vintage year (2024)",
  "ind_t211_2024_cycle",              "Projection cycle (2024-2034)",
  "ind_t211_2024_total_pct",          "Total economy projected growth %",
  "ind_t211_2024_total_cagr",         "Total economy CAGR %",
  "ind_t211_2024_hc_pct",             "Healthcare sector projected growth %",
  "ind_t211_2024_hc_cagr",            "Healthcare sector CAGR %",
  "ind_t211_2024_hc_emp",             "Healthcare employment, base year (thousands)",
  "ind_t211_2024_shrs_ind",           "SHRS-relevant industries dataframe with growth metrics",
  "ind_t211_2024_ind_621340",         "NAICS 621340 row (PT/OT/SLP/AuD offices)",
  "ind_t211_2024_ind_621990",         "NAICS 621990 row (All other ambulatory HC)",
  "ind_t211_2024_program_scores",     "Program-to-industry scorecard (9 programs, 0-100 composite)",
  "ind_t211_2024_hc_subsectors",      "HC sub-sector composition (ambulatory, hospitals, nursing, social)",
  "ind_t211_2024_full_df",            "Complete Table 2.11 dataframe (all industries)"
)

output_registry %>%
  kable(caption = "Output Variable Registry — industry_t211_2024.Rmd") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
