---
title: "SHRS Competitor Analysis - Competitor Identification Report"
author: "SHRS MQE Capstone Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r comp-rpt-setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# 0 - About This Report

This report reads the CSVs produced by `competitor_01_pull.R` and presents
the competitor landscape for all 9 SHRS programs. No API calls are made here.

```{r comp-rpt-packages}
library(tidyverse)
library(scales)
library(kableExtra)
```

```{r comp-rpt-load}
# -- EDIT THIS PATH if your CSVs are in a different location --
DATA_DIR <- "competitor_data"

tier1          <- read_csv(file.path(DATA_DIR, "tier1_statewide.csv"), show_col_types = FALSE)
tier2          <- read_csv(file.path(DATA_DIR, "tier2_regional.csv"), show_col_types = FALSE)
tier1_summary  <- read_csv(file.path(DATA_DIR, "tier1_summary.csv"), show_col_types = FALSE)
tier2_summary  <- read_csv(file.path(DATA_DIR, "tier2_summary.csv"), show_col_types = FALSE)
master_roster  <- read_csv(file.path(DATA_DIR, "master_roster.csv"), show_col_types = FALSE)
pitt_rank      <- read_csv(file.path(DATA_DIR, "pitt_rank.csv"), show_col_types = FALSE)
cip_crosswalk  <- read_csv(file.path(DATA_DIR, "cip_crosswalk.csv"), show_col_types = FALSE)

PITT_UNITID <- 215293L
latest_year <- max(tier1$year, na.rm = TRUE)
if (is.na(latest_year) || is.infinite(latest_year)) {
  latest_year <- 2022
}

cat("Data loaded from:", DATA_DIR, "\n")
cat("Programs:", paste(sort(unique(tier1$shrs_program)), collapse = ", "), "\n")
cat("Tier 1 rows:", nrow(tier1), "| Tier 2 rows:", nrow(tier2), "\n")
```

---

# 1 - Tier 1: Statewide Competition (All PA)

How many PA institutions compete with Pitt in each program, and what is
Pitt's share of statewide graduate production?

## 1.1 Summary

```{r comp-rpt-t1-summary}
tier1_summary |>
  arrange(desc(total_pa_graduates)) |>
  select(
    Program          = shrs_program,
    Dept             = shrs_dept,
    `PA Schools`     = n_competitors,
    `Competitors`    = n_competitors_excl,
    `Total PA Grads` = total_pa_graduates,
    `Pitt Grads`     = pitt_graduates,
    `Pitt Share (%)`  = pitt_share_pct
  ) |>
  kable(format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 1.2 Pitt's Share Visualization

```{r comp-rpt-t1-share, fig.height=5}
tier1_summary |>
  mutate(other_graduates = total_pa_graduates - pitt_graduates) |>
  select(shrs_program, Pitt = pitt_graduates, `Other PA` = other_graduates) |>
  pivot_longer(-shrs_program, names_to = "source", values_to = "grads") |>
  ggplot(aes(x = reorder(shrs_program, grads, sum), y = grads, fill = source)) +
  geom_col(width = 0.6) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("Pitt" = "#003594", "Other PA" = "#CCCCCC")) +
  labs(
    title = "Pitt vs. All Other PA Institutions: Graduate Production",
    subtitle = paste0("IPEDS Completions, ", latest_year),
    x = NULL, y = "Total Degrees Awarded", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## 1.3 Pitt's Rank in PA

```{r comp-rpt-pitt-rank}
pitt_rank |>
  arrange(rank) |>
  select(
    Program     = shrs_program,
    `Pitt Awards` = total_awards,
    `Rank in PA`  = rank,
    `of N Schools` = n_schools,
    Percentile    = percentile
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 1.4 Full Competitor Lists by Program

```{r comp-rpt-t1-lists, results='asis', fig.height=6}
for (prog in sort(unique(tier1$shrs_program))) {

  cat(paste0("\n### ", prog, "\n\n"))

  prog_data <- tier1 |>
    filter(shrs_program == prog) |>
    mutate(is_pitt = unitid == PITT_UNITID)

  # Table
  prog_data |>
    mutate(Institution = if_else(is_pitt, paste0("**", inst_name, " (PITT)**"), inst_name)) |>
    select(Institution, City = city, Awards = total_awards, `Degree Levels` = degree_levels) |>
    kable(format.args = list(big.mark = ",")) |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
    print()

  # Chart
  p <- prog_data |>
    ggplot(aes(x = reorder(inst_name, total_awards), y = total_awards, fill = is_pitt)) +
    geom_col(width = 0.6) +
    coord_flip() +
    scale_fill_manual(values = c("FALSE" = "#CCCCCC", "TRUE" = "#003594"),
                      labels = c("Competitor", "Pitt"), guide = "none") +
    scale_y_continuous(labels = comma) +
    labs(title = paste0(prog, ": PA Competitor Landscape"),
         x = NULL, y = "Degrees Awarded") +
    theme_minimal(base_size = 11)
  print(p)

  cat("\n\n")
}
```

---

# 2 - Tier 2: Regional Competition (Pittsburgh-Erie Corridor)

The Pittsburgh-Erie corridor covers 17 counties in western and northwestern PA.
These are the institutions a local prospective student would most directly
weigh against Pitt.

## 2.1 Summary

```{r comp-rpt-t2-summary}
tier2_summary |>
  arrange(desc(total_grads)) |>
  select(
    Program              = shrs_program,
    `Regional Competitors` = n_competitors,
    `Regional Grads`       = total_grads,
    `Pitt Grads`           = pitt_grads,
    `Pitt Share (%)`       = pitt_share_pct
  ) |>
  kable(format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 2.2 Regional Share Visualization

```{r comp-rpt-t2-share, fig.height=5}
tier2_summary |>
  mutate(other_grads = total_grads - pitt_grads) |>
  select(shrs_program, Pitt = pitt_grads, `Other Regional` = other_grads) |>
  pivot_longer(-shrs_program, names_to = "source", values_to = "grads") |>
  ggplot(aes(x = reorder(shrs_program, grads, sum), y = grads, fill = source)) +
  geom_col(width = 0.6) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("Pitt" = "#003594", "Other Regional" = "#FDAE61")) +
  labs(
    title = "Pitt vs. Regional Competitors: Graduate Production",
    subtitle = "Pittsburgh-Erie Corridor (17 counties)",
    x = NULL, y = "Total Degrees Awarded", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## 2.3 Regional Competitor Lists by Program

```{r comp-rpt-t2-lists, results='asis', fig.height=5}
for (prog in sort(unique(tier2$shrs_program))) {

  cat(paste0("\n### ", prog, "\n\n"))

  prog_data <- tier2 |>
    filter(shrs_program == prog) |>
    mutate(is_pitt = unitid == PITT_UNITID)

  prog_data |>
    mutate(Institution = if_else(is_pitt, paste0("**", inst_name, " (PITT)**"), inst_name)) |>
    select(Institution, City = city, Region = region, Awards = total_awards,
           `Degree Levels` = degree_levels) |>
    kable(format.args = list(big.mark = ",")) |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
    print()

  cat("\n")
}
```

---

# 3 - Cross-Program Analysis

## 3.1 Most Frequent Competitors

Which PA institutions compete with Pitt across the most programs?

```{r comp-rpt-frequent}
master_roster |>
  filter(!is_pitt) |>
  group_by(unitid, inst_name, city) |>
  summarise(
    n_programs   = n(),
    programs     = paste(sort(unique(shrs_program)), collapse = ", "),
    total_awards = sum(total_awards, na.rm = TRUE),
    .groups      = "drop"
  ) |>
  arrange(desc(n_programs), desc(total_awards)) |>
  head(20) |>
  select(
    Institution  = inst_name,
    City         = city,
    `# Programs` = n_programs,
    Programs     = programs,
    `Total Awards` = total_awards
  ) |>
  kable(format.args = list(big.mark = ",")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 3.2 Competition Intensity by Program

How crowded is each program's competitive landscape?

```{r comp-rpt-intensity, fig.height=5}
tier1_summary |>
  ggplot(aes(x = reorder(shrs_program, n_competitors),
             y = n_competitors, fill = shrs_dept)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n_competitors), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Number of PA Competitors by SHRS Program",
    subtitle = "Includes all institutions awarding degrees in matching CIP codes",
    x = NULL, y = "Number of PA Institutions", fill = "Department"
  ) +
  theme_minimal(base_size = 13)
```

## 3.3 Pitt's Position: Percentile by Program

```{r comp-rpt-percentile, fig.height=5}
pitt_rank |>
  ggplot(aes(x = reorder(shrs_program, percentile), y = percentile,
             fill = percentile)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(percentile, "%")), hjust = -0.1, size = 4) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "gray50") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 110)) +
  scale_fill_gradient2(low = "#d7191c", mid = "#fee08b", high = "#1a9850",
                       midpoint = 50, guide = "none") +
  labs(
    title = "Pitt's Percentile Rank in PA by Program",
    subtitle = "Based on completions volume. 100% = largest producer in PA.",
    x = NULL, y = "Percentile"
  ) +
  theme_minimal(base_size = 13)
```

---

# 4 - Tier 1 vs. Tier 2 Comparison

How does the competitive picture change when we zoom from statewide to regional?

```{r comp-rpt-tier-compare}
comparison <- tier1_summary |>
  select(shrs_program,
         t1_schools = n_competitors,
         t1_grads   = total_pa_graduates,
         t1_pitt    = pitt_share_pct) |>
  left_join(
    tier2_summary |>
      select(shrs_program,
             t2_schools = n_competitors,
             t2_grads   = total_grads,
             t2_pitt    = pitt_share_pct),
    by = "shrs_program"
  ) |>
  mutate(
    share_increase = round(t2_pitt - t1_pitt, 1)
  )

comparison |>
  select(
    Program               = shrs_program,
    `PA Schools`          = t1_schools,
    `Regional Schools`    = t2_schools,
    `Pitt Share PA (%)`   = t1_pitt,
    `Pitt Share Reg (%)` = t2_pitt,
    `Share Increase (pp)` = share_increase
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r comp-rpt-tier-compare-chart, fig.height=5}
comparison |>
  select(shrs_program, `Statewide` = t1_pitt, `Regional` = t2_pitt) |>
  pivot_longer(-shrs_program, names_to = "scope", values_to = "pitt_share") |>
  ggplot(aes(x = reorder(shrs_program, pitt_share, max),
             y = pitt_share, fill = scope)) +
  geom_col(position = "dodge", width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c("Statewide" = "#2c7bb6", "Regional" = "#003594")) +
  labs(
    title = "Pitt's Market Share: Statewide vs. Regional",
    subtitle = "Share increases when zooming to the Pittsburgh-Erie corridor",
    x = NULL, y = "Pitt Share of Graduates (%)", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

# 5 - Key Takeaways

```{r comp-rpt-takeaways, results='asis'}
cat("## Headline Findings\n\n")

# Strongest positions
strong <- pitt_rank |> filter(percentile >= 75) |> arrange(desc(percentile))
if (nrow(strong) > 0) {
  cat("**Strongest competitive positions** (75th+ percentile in PA):\n\n")
  for (i in seq_len(nrow(strong))) {
    cat(paste0("- **", strong$shrs_program[i], "**: Rank ",
               strong$rank[i], " of ", strong$n_schools[i],
               " (", strong$percentile[i], "th percentile, ",
               strong$total_awards[i], " awards)\n"))
  }
  cat("\n")
}

# Weaker positions
weak <- pitt_rank |> filter(percentile < 50) |> arrange(percentile)
if (nrow(weak) > 0) {
  cat("**Areas for attention** (below 50th percentile in PA):\n\n")
  for (i in seq_len(nrow(weak))) {
    cat(paste0("- **", weak$shrs_program[i], "**: Rank ",
               weak$rank[i], " of ", weak$n_schools[i],
               " (", weak$percentile[i], "th percentile, ",
               weak$total_awards[i], " awards)\n"))
  }
  cat("\n")
}

# Missing from ranking
missing <- setdiff(sort(unique(tier1$shrs_program)), pitt_rank$shrs_program)
if (length(missing) > 0) {
  cat("**Programs where Pitt shows 0 IPEDS completions**: ",
      paste(missing, collapse = ", "),
      ". These may be coded under different CIP codes in IPEDS or may not currently be awarding degrees.\n\n")
}

# Most competitive markets
crowded <- tier1_summary |> arrange(desc(n_competitors)) |> head(3)
cat("**Most crowded PA markets**: ",
    paste0(crowded$shrs_program, " (", crowded$n_competitors, " schools)"),
    sep = "", collapse = ", ")
cat("\n\n")

# Regional dominance
reg_strong <- tier2_summary |> filter(pitt_share_pct >= 30) |> arrange(desc(pitt_share_pct))
if (nrow(reg_strong) > 0) {
  cat("**Regional dominance** (30%+ share in Pittsburgh-Erie corridor):\n\n")
  for (i in seq_len(nrow(reg_strong))) {
    cat(paste0("- **", reg_strong$shrs_program[i], "**: ",
               reg_strong$pitt_share_pct[i], "% regional share\n"))
  }
}
```

---

# Appendix

**Data Source**: IPEDS Completions by 6-digit CIP code, accessed via Urban
Institute Education Data Portal API. Institution directory from the same source.

**Tier Definitions**:

- Tier 1 (Statewide): All PA institutions with non-zero completions in target CIP codes
- Tier 2 (Regional): Tier 1 subset in 17 counties from Pittsburgh metro through Erie
- Tier 3 (National): U.S. News top programs (not yet populated)

**Latest IPEDS Year**: `r latest_year`

**Note on AT and DN**: Pitt may show 0 completions for Athletic Training and
Dietetics/Nutrition if these programs are coded under different CIP codes in
Pitt's IPEDS reporting. Verify with SHRS registrar.
