---
title: "SHRS Competitor Analysis - Tuition Comparison Report"
author: "SHRS MQE Capstone Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r comp-tui-setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# 0 - About This Report

This report reads the CSVs produced by `competitor_02_tuition_pull.R` and
presents Pitt's tuition positioning against PA competitors for all 9 SHRS
programs. No API calls are made here.

```{r comp-tui-load}
library(tidyverse)
library(scales)
library(kableExtra)

DATA_DIR <- "competitor_data"

tier1_tuition      <- read_csv(file.path(DATA_DIR, "tier1_with_tuition.csv"), show_col_types = FALSE)
tier2_tuition      <- read_csv(file.path(DATA_DIR, "tier2_with_tuition.csv"), show_col_types = FALSE)
tuition_comparison <- read_csv(file.path(DATA_DIR, "tuition_comparison.csv"), show_col_types = FALSE)
tuition_latest     <- read_csv(file.path(DATA_DIR, "tuition_latest.csv"), show_col_types = FALSE)

PITT_UNITID <- 215293L

pitt_tuition <- tuition_latest |> filter(unitid == PITT_UNITID)
pitt_in  <- pitt_tuition$tuition_in_state[1]
pitt_out <- pitt_tuition$tuition_out_of_state[1]

cat("Pitt graduate tuition (in-state):", dollar(pitt_in), "\n")
cat("Pitt graduate tuition (out-of-state):", dollar(pitt_out), "\n")
```

---

# 1 - Pitt vs. PA Competitors: Summary

How does Pitt's graduate tuition compare to the average and median of
competitors for each program?

```{r comp-tui-summary-table}
tuition_comparison |>
  arrange(desc(pitt_vs_avg_in)) |>
  mutate(
    pitt_in_state     = dollar(pitt_in_state),
    pitt_out_of_state = dollar(pitt_out_of_state),
    avg_in_state      = dollar(avg_in_state),
    med_in_state      = dollar(med_in_state),
    min_in_state      = dollar(min_in_state),
    max_in_state      = dollar(max_in_state),
    pitt_vs_avg_in    = dollar(pitt_vs_avg_in)
  ) |>
  select(
    Program            = shrs_program,
    `Pitt In-State`    = pitt_in_state,
    `Avg Competitor`   = avg_in_state,
    `Median Comp`      = med_in_state,
    `Min`              = min_in_state,
    `Max`              = max_in_state,
    `Pitt vs Avg`      = pitt_vs_avg_in
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r comp-tui-summary-chart, fig.height=6}
tuition_comparison |>
  filter(!is.na(pitt_in_state)) |>
  select(shrs_program, Pitt = pitt_in_state, `Competitor Avg` = avg_in_state,
         `Competitor Med` = med_in_state) |>
  pivot_longer(-shrs_program, names_to = "measure", values_to = "tuition") |>
  ggplot(aes(x = reorder(shrs_program, tuition, max),
             y = tuition, fill = measure)) +
  geom_col(position = "dodge", width = 0.6) +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  scale_fill_manual(values = c(
    "Pitt" = "#003594",
    "Competitor Avg" = "#CCCCCC",
    "Competitor Med" = "#999999"
  )) +
  labs(
    title = "Pitt Graduate Tuition vs. PA Competitors (In-State)",
    subtitle = "IPEDS Academic Year Tuition, 2021",
    x = NULL, y = "Graduate Tuition & Fees (Full-Time)", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

# 2 - Tuition Range by Program

Where does Pitt fall within the full range of competitor tuition?

```{r comp-tui-range, fig.height=7}
tier1_tuition |>
  filter(!is.na(tuition_in_state)) |>
  ggplot(aes(x = reorder(shrs_program, tuition_in_state, median),
             y = tuition_in_state)) +
  geom_boxplot(fill = "#E8E8E8", outlier.shape = NA, width = 0.5) +
  geom_jitter(aes(color = is_pitt, size = is_pitt),
              width = 0.15, alpha = 0.7) +
  scale_y_continuous(labels = dollar) +
  scale_color_manual(values = c("FALSE" = "#999999", "TRUE" = "#003594"),
                     labels = c("Competitor", "Pitt")) +
  scale_size_manual(values = c("FALSE" = 2, "TRUE" = 5),
                    labels = c("Competitor", "Pitt")) +
  coord_flip() +
  labs(
    title = "Graduate Tuition Distribution by SHRS Program",
    subtitle = "Pitt highlighted in blue. Box = IQR of PA competitors.",
    x = NULL, y = "In-State Graduate Tuition & Fees",
    color = NULL, size = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

# 3 - Detailed Tuition by Program

```{r comp-tui-detail, results='asis', fig.height=5}
for (prog in sort(unique(tier1_tuition$shrs_program))) {

  prog_data <- tier1_tuition |>
    filter(shrs_program == prog, !is.na(tuition_in_state)) |>
    arrange(tuition_in_state)

  if (nrow(prog_data) == 0) next

  cat(paste0("\n## ", prog, "\n\n"))

  # Table
  prog_data |>
    mutate(
      Institution = if_else(is_pitt, paste0("**", inst_name, " (PITT)**"), inst_name),
      tuition_in_state = dollar(tuition_in_state),
      tuition_out_of_state = dollar(tuition_out_of_state)
    ) |>
    select(
      Institution,
      City = city,
      `In-State` = tuition_in_state,
      `Out-of-State` = tuition_out_of_state,
      `Awards` = total_awards
    ) |>
    kable() |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
    print()

  # Chart
  p <- prog_data |>
    ggplot(aes(x = reorder(inst_name, tuition_in_state),
               y = tuition_in_state, fill = is_pitt)) +
    geom_col(width = 0.6) +
    coord_flip() +
    scale_y_continuous(labels = dollar) +
    scale_fill_manual(values = c("FALSE" = "#CCCCCC", "TRUE" = "#003594"),
                      guide = "none") +
    labs(title = paste0(prog, ": In-State Graduate Tuition"),
         x = NULL, y = "Tuition & Fees") +
    theme_minimal(base_size = 11)
  print(p)

  cat("\n\n")
}
```

---

# 4 - Regional Tuition Comparison (Pittsburgh-Erie)

How does Pitt compare to just the local competitors?

```{r comp-tui-regional, fig.height=6}
tier2_tuition |>
  filter(!is.na(tuition_in_state)) |>
  ggplot(aes(x = reorder(inst_name, tuition_in_state),
             y = tuition_in_state, fill = is_pitt)) +
  geom_col(width = 0.6) +
  coord_flip() +
  facet_wrap(~ shrs_program, scales = "free", ncol = 2) +
  scale_y_continuous(labels = dollar) +
  scale_fill_manual(values = c("FALSE" = "#FDAE61", "TRUE" = "#003594"),
                    labels = c("Regional Competitor", "Pitt")) +
  labs(
    title = "Regional Tuition Comparison: Pitt vs Pittsburgh-Erie Corridor",
    subtitle = "Graduate in-state tuition & fees",
    x = NULL, y = "Tuition & Fees", fill = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_text(size = 7))
```

---

# 5 - In-State vs Out-of-State Spread

How much more do out-of-state students pay? This matters for understanding
cost competitiveness for students crossing state lines.

```{r comp-tui-spread, fig.height=6}
tier1_tuition |>
  filter(!is.na(tuition_in_state), !is.na(tuition_out_of_state)) |>
  mutate(spread = tuition_out_of_state - tuition_in_state) |>
  ggplot(aes(x = tuition_in_state, y = tuition_out_of_state,
             color = is_pitt, size = is_pitt)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  scale_color_manual(values = c("FALSE" = "#999999", "TRUE" = "#003594"),
                     labels = c("Competitor", "Pitt")) +
  scale_size_manual(values = c("FALSE" = 2, "TRUE" = 5),
                    labels = c("Competitor", "Pitt")) +
  labs(
    title = "In-State vs Out-of-State Graduate Tuition",
    subtitle = "Points above the dashed line charge more for out-of-state. Pitt in blue.",
    x = "In-State Tuition", y = "Out-of-State Tuition",
    color = NULL, size = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

```{r comp-tui-spread-table}
tier1_tuition |>
  filter(!is.na(tuition_in_state), !is.na(tuition_out_of_state)) |>
  mutate(
    spread     = tuition_out_of_state - tuition_in_state,
    spread_pct = round(spread / tuition_in_state * 100, 0)
  ) |>
  group_by(shrs_program) |>
  summarise(
    pitt_spread     = dollar(max(spread[is_pitt], na.rm = TRUE)),
    pitt_spread_pct = paste0(max(spread_pct[is_pitt], na.rm = TRUE), "%"),
    avg_spread      = dollar(round(mean(spread[!is_pitt], na.rm = TRUE))),
    avg_spread_pct  = paste0(round(mean(spread_pct[!is_pitt], na.rm = TRUE)), "%"),
    .groups = "drop"
  ) |>
  select(
    Program           = shrs_program,
    `Pitt Spread`     = pitt_spread,
    `Pitt Spread %`   = pitt_spread_pct,
    `Avg Comp Spread`  = avg_spread,
    `Avg Comp Spread %` = avg_spread_pct
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 6 - Key Findings

```{r comp-tui-findings, results='asis'}
# Programs where Pitt is more expensive than average
expensive <- tuition_comparison |>
  filter(!is.na(pitt_vs_avg_in), pitt_vs_avg_in > 0) |>
  arrange(desc(pitt_vs_avg_in))

if (nrow(expensive) > 0) {
  cat("## Pitt is above average in-state tuition for:\n\n")
  for (i in seq_len(nrow(expensive))) {
    cat(paste0("- **", expensive$shrs_program[i], "**: ",
               dollar(expensive$pitt_in_state[i]),
               " vs ", dollar(expensive$avg_in_state[i]),
               " avg (+", dollar(expensive$pitt_vs_avg_in[i]), ")\n"))
  }
  cat("\n")
}

# Programs where Pitt is cheaper
cheaper <- tuition_comparison |>
  filter(!is.na(pitt_vs_avg_in), pitt_vs_avg_in < 0) |>
  arrange(pitt_vs_avg_in)

if (nrow(cheaper) > 0) {
  cat("## Pitt is below average in-state tuition for:\n\n")
  for (i in seq_len(nrow(cheaper))) {
    cat(paste0("- **", cheaper$shrs_program[i], "**: ",
               dollar(cheaper$pitt_in_state[i]),
               " vs ", dollar(cheaper$avg_in_state[i]),
               " avg (", dollar(cheaper$pitt_vs_avg_in[i]), ")\n"))
  }
  cat("\n")
}

# Programs with no Pitt tuition data
missing <- tuition_comparison |> filter(is.na(pitt_in_state))
if (nrow(missing) > 0) {
  cat("## No Pitt tuition comparison available for: ",
      paste(missing$shrs_program, collapse = ", "),
      "\n\n(Pitt shows 0 IPEDS completions in these CIP codes.)\n\n")
}

cat("## Context\n\n")
cat(paste0("- IPEDS tuition is **institution-level**, not program-specific. ",
           "Actual program costs may differ due to differential tuition, fees, or clinical charges.\n"))
cat(paste0("- Data is from **academic year 2021** (most recent available in IPEDS).\n"))
cat(paste0("- Pitt graduate in-state: **", dollar(pitt_in), "** | ",
           "Out-of-state: **", dollar(pitt_out), "**\n"))
```

---

# Appendix

**Data Source**: IPEDS Academic Year Tuition via Urban Institute Education Data Portal API.

**Tuition Type Codes**: 2 = in-district, 3 = in-state, 4 = out-of-state.
We use type 3 (in-state) and type 4 (out-of-state) for graduate level (level_of_study = 2).

**Important caveat**: IPEDS reports institution-level tuition, not program-specific.
Many health professions programs charge differential tuition above the base rate.
The numbers here represent the institutional baseline. Actual program costs for
clinical doctorates (DPT, OTD, AuD) and professional programs (PAS) are typically
higher than the institutional rate shown.
